---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{get_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: 
  - "Andreas G. Damsbo"
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = TRUE,
  comment = "#>"
)
```


```{r setup, include=FALSE}
# Packages needed only for the vignette (they are listed in Suggests:)
library(lubridate)  
library(ggplot2)    
library(dahcalc)    

# Make the vignette reproducible
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
```

## What the DAH pipeline does  

The **DAH (Days‑At‑Home)** pipeline takes a *long‑format* table of clinical events and calculates, for each patient, how many days out of a user‑defined window (default 30 days) they spent **at home**.  

*Primary admission* → start of the window.  
*Hospital* or *rehabilitation* stays → days **not** at home.  
If a patient dies **inside** the window, their DAH is set to **0**.  

All column names can be customised; you only need to tell the wrapper which columns correspond to the required concepts.

---

## 1️⃣ Get the package and load the example data set  

```{r eval=FALSE}
pak::pak("agdamsbo/dahcalc")
library(dahcalc)
```


The package ships a reproducible synthetic data set called `synthetic_data`.

```{r}
str(dahcalc::synthetic_data)
```

The data already have the **canonical column names** that the pipeline expects:

* `patient_id`
* `event_id`
* `event_type` (`"primary"`, `"hospital"`, `"rehabilitation"` or `"death"`)
* `start_date`, `end_date`
* `intervention_date` (only filled for the primary admission)
* `death_date` (filled only for rows of type `"death"`)

---

## 2️⃣ (Optional) Rename columns to match your own file  

If your own CSV uses different names you can rename them before calling the
pipeline.  For demonstration we rename them to a lower‑case scheme, but you
could skip this step entirely if your file already uses the canonical names.

```{r}
# Make a copy so we do not alter the original object
my_dt <- dahcalc::synthetic_data

# Suppose the user’s file uses the following names:
#  pid, eid, etype, adm_start, adm_end, int_date, dod
my_dt <- setNames(
  my_dt,
  c("pid", "eid", "etype", "adm_start",
          "adm_end", "int_date", "dod")
)

head(my_dt)
```

Now `my_dt` looks like the data a user might supply.

---

## 3️⃣ Run the DAH pipeline  

Tell the wrapper the names you are using (or skip the arguments if you kept
the canonical names).  

```{r}
res <- run_dah_pipeline(
  data_long             = my_dt,
  window_days           = 30,        # 30‑day observation window
  patient_id_col        = "pid",
  event_id_col          = "eid",
  event_type_col        = "etype",
  start_date_col        = "adm_start",
  end_date_col          = "adm_end",
  intervention_date_col = "int_date",
  death_date_col        = "dod",
  verbose               = TRUE,      # prints a short validation summary
  keep_original_names   = TRUE        # keep the user’s column names in the output
)

# The result is a list with three main components:
str(res, max.level = 1)
```

* `res$per_patient` – one row per patient with DAH values.  
* `res$cohort_summary` – aggregated statistics for the whole cohort.  
* `res$plot` – a ggplot object showing the DAH distribution.

---

## 4️⃣ Look at the cohort‑level summary  

```{r}
print(res$cohort_summary)
```

Typical output shows:

* number of patients,
* mean / median / SD of DAH,
* percentage with **full** 30 days at home,
* DAH per 100 person‑days, etc.

---

## 5️⃣ Inspect the per‑patient table  

```{r}
head(res$per_patient, 10)
```

Key columns you’ll see:

| column | meaning |
|--------|---------|
| `pid` (or `patient_id`) | patient identifier |
| `dah` | days at home within the window (0 if death occurs inside) |
| `effective_window` | length of the observation window actually used (shorter when death occurs) |
| `institutional_days` | number of days spent in hospital/rehab |
| `died_in_window` | `TRUE` if death happened inside the window |

---

## 6️⃣ Visualise the DAH distribution  

```{r}
print(res$plot)   # displays a histogram with a vertical line at 30 days
```

The histogram lets you quickly see how many patients spent all 30 days at
home, how many had long institutional stays, and where deaths fall.

---

## 7️⃣ What to do with your own data  

When you have a real CSV file, the workflow is the same:

```{r eval=FALSE}
my_real_data <- read.csv("my_clinical_file.csv")

# (optional) rename columns if they differ from the canonical ones
# setnames(my_real_data, old = "...", new = "...")

result <- run_dah_pipeline(
  dt_long               = my_real_data,
  patient_id_col        = "my_id",
  event_id_col          = "my_event_id",
  event_type_col        = "my_type",
  start_date_col        = "my_start",
  end_date_col          = "my_end",
  intervention_date_col = "my_intervention",
  death_date_col        = "my_death",
  window_days           = 30,
  keep_original_names   = TRUE
)

# Then explore result$cohort_summary, result$per_patient, result$plot …
```

That’s all you need to obtain **Days‑At‑Home** metrics for any data set that follows the required event structure.
